The \tsc{r}-pronoun and postposition \tit{waarmee} in Dutch

### 2020-01-16 paper.tex
klaar, alleen nog lezen
pfoe, wat een bevalling

Dutch has the preposition \tit{met} `with' that expresses instrumental case. In \ref{ex:metdp}, \tit{met} `with' is combined with a full DP. The inanimate pronoun in Dutch is \tit{'t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{'t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{'t} `it' can be used as the object of a verb.

@@

\tit{Met} `with' and \tit{'t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{'r} and the postposition \tit{mee}, as shown in \ref{ex:jarmee}.\footnote{The \tsc{r}-pronoun \tit{'r} in \ref{ex:klimerop} can be written as \tit{er}, \tit{der} and \tit{'r} and pronounced as respectively /ɛr/, /dər/ or /ər/. As far as I am aware, there is no clear meaning difference between these forms. See \citet{wesseling2018} for discussion.
In my examples I use \tit{'r}, but the other two forms fit just as well.}$^,$\footnote{In this paper I do not make any claims about the distinction between prefixes and prepositions, or suffixes and postpositions.}

@@

\tsc{R}-pronouns \citep{riemsdijk1978,koopman1994} are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r}. The adposition it combines with obligatorily follows the \tsc{r}-pronoun, compare \ref{ex:jarmee} and \ref{ex:neemeer}.
Notice also that the preposition \tit{met} `with' has changed form into \tit{mee} `with' when it appears as a postposition (see \ref{ex:metdp} and \ref{ex:jarmee}).

@@

This chapter is concerned with the question of how to correctly rule out the construction in \ref{ex:neemett} and to let the \tsc{r}-pronoun construction in \ref{ex:jarmee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout. Just like \citet{riemsdijk1978} I analyze \tsc{r}-pronouns and postpositions as a type of allomorphy. What is different is that spellout rules out the ungrammaticality of the preposition and inanimate pronoun, rather than the stipulation of a filter.

This chapter is an in-depth study of the instrumental \tsc{r}-pronoun-postposition combination \tit{'r -mee} `with it' and \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronoun-postpositions combinations, the meaning component of the locative is intuitive, as many prepositions express locations, directions etc. However, an instrumental expresses an instrument and does not have a meaning component assosicated with location.
The second reason why I focus on this particular \tsc{r}-pronoun and postposition has to do with the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun.\footnote{I do not have anything to say about the distinction between prefixes and prepositions, or suffixes and postpositions.} This last observervation has so far remained unexplained.

%%here the basic idea%%
When all features form a proper constituent, then the \tsc{r}-pronoun surfaces. When there is a complement or whatever, it does not work.
In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a proper constituent (i.e. a constituent to the exclusion of other features), the \tsc{r}-pronoun surfaces. If it is not a proper constituent, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. --here more about the internal structure of r-pronouns, and the postposition preposition distinction--

%%here the structure of the paper%%
This paper is structured as follows. In Section \ref{sec:distribution} I discuss the distribution of \tit{waarmee} `waarmee' and \tit{met wat} `with what'. I show that the \tsc{r}-pronoun surfaces when it forms a proper constituent, and \tit{met wat} `with what' when it does not. I decompose \tit{waarmee} `with what' and \tit{met wat} `with what' to make sense of the phonological similarities, and I provide a more detailed analysis. Unmarked examples are constructed and have been verified by native speakers.

@@

@@

The goal of this section is to show that \tit{waarmee} `with what' is the default as instrumental relative pronoun, a shown in a.o. \citet{riemsdijk1978,koopman2000}. In order to show that \tit{waarmee} `with what' is the default, I discuss the distribution of \tsc{r}-pronouns and regular pronouns in more general. I start with the personal pronouns and then return to the wh-pronouns.

Dutch has the personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{het} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

@@

The example in \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. However, the inanimate personal pronoun \tit{het} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an ʀ-pronoun appears. This is illustrated in \ref{ex:preper}. \ref{ex:erprep} shows that the \tsc{r}-pronoun obligatorily moves to the left of the pronoun.

@@

\tit{Met} is not the only preposition with which this happens. \tit{Op} `on' and \tit{in} `in' do not combine with the inanimate personal pronoun \tit{'t}, but the \tsc{r}-pronoun is used obligatorily.

@@

The situation of the inanimate wh-pronouns resembles the inanimate personal pronouns. \tit{Wat} `what' can function as an object of a verb (see \ref{ex:wat}), but not as an object of a preposition \ref{ex:metwat}. In that case, the \tsc{r}-pronoun \tit{waarmee} `with what' is appears, as shown in \ref{ex:waarmee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes is only acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice for the object hearer is painting with.}

@@

\tit{'t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by respectively \tit{'r} `there' and \tit{waar} `where'.

The next section discusses the role of constituency in \tsc{r}-pronouns.

@@

Sometimes \tit{met wat} shows up!

Evidence from that comes from the complementary distribution of \tit{waarmee} and \tit{met wat} in mismatching free relatives.

This is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I illustrate this in \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative DP. The \tsc{r}-pronoun \tit{waarmee} `with what' is used here.\footnote{In this example, \tit{waar} `where' takes \tit{-mee} `with' to the left edge of the embedded clause. It is also possible for \tit{-mee} `with' to be stranded, and \tit{waar} `where' to be moved to the left edge of the embedded clause on its own.

@@

\exg. Ik heb gekocht waarmee jij schildert.\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}

If the predicates are switched around between the clauses, the \tsc{r}-pronoun does not appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental object in the main clause and \tit{gekocht} `bought' combines with an accusative object in the embedded clause. The use of an \tsc{r}-pronoun is ungrammatical, as indicated by the ungrammaticality of \ref{ex:schilderwaarmee}. Instead, a combination of the regular instrumental preposition \tit{met} `with' and the regular wh-pronoun \tit{was} `what' in used.

@@

The use of \tit{met wat} `with what' is ungrammatical in the context in which \tit{waarmee} `with what' appeared in \ref{ex:gekochtwaarmee}. This is illustrated in \ref{ex:gekochtmetwat}.

\exg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

In the introduction I discussed the distribution between \tit{waarmee} `with what' and \tit{met wat} `with what' in free relatives with predicates that combine with different cases. Table \ref{tbl:distribution} repeats the generalization. When the main clause predicate combines with an accusative and the embedded clause predicate with an instrumental, \tit{waarmee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the embedded clause predicate with an accusative, \tit{waarmee} `with what' is ungrammatical and \tit{met wat} `with what' is used.

@@

In this section I first show that \tsc{r}-pronouns are the default complement of a preposition. Next, I illustrate that a necessary requirement for an \tsc{r}-pronoun is that is forms a proper constituent.

Let me now return to the mismatching free relatives. I repeat the relevant grammatical examples in \ref{ex:grammatical}.

\ex.\label{ex:grammatical}
\ag. Ik heb gekocht waarmee jij schildert.\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:grammaticalwaar}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:grammaticalwat}

In the preceding section I showed that ʀ-pronouns are expected in combinations with prepositions. This means that the use of \tit{waarmee} `with what' \ref{ex:grammaticalwaar} is not surprising. Something that is surprising is the use of \tit{met wat} `with what' in \ref{ex:grammaticalwat}, and this is the example something more needs to be said about. In the remainder of this section I argue that this `something more' is that the instrumental object in \ref{ex:grammaticalwat} does not form a proper constituent, i.e. it is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns contain a object that does form a proper constituent.

@@

In each of these examples the instrumental object forms a constituent at a certain point in the derivation. In \ref{ex:const1}, the instrumental object forms a proper constituent in the surface order, as shown in \ref{ex:const1stage}. In \ref{ex:const2}, the instrumental object forms a proper constituent before wh- and V2- movement, shown in \ref{ex:const2stage}.
The structure in \ref{ex:const3stage} represents a stage in the derivation of the embedded clauses in \ref{ex:const3} and \ref{ex:const4}. Again, in the stage, which comes before relative movement of the pronoun to the left periphery of the relative clause, the instrumental object forms a proper constituent.

@@

%There is additional evidence for the fact that \tit{waarmee} `with what' forms a constituent in the constructions above. In the example in \ref{ex:const2}-\ref{ex:const4} only \tit{waar} `where' was fronted, but \ref{ex:constwaarmee} shows that also the phrase containing \tit{-mee} with' can be moved. As wh-movement can only target constituents, it follows that \tit{waar} `where' and \tit{-mee} `with' have to form a constituent.

%\ex.\label{ex:constwaarmee}
%\ag. Waar -mee schilder jij?\\
 %{where -with} paint you\\
 %`What are you painting with?'
%\bg. Ik schilder [waar -mee jij ook schildert].\\
 %I paint {where -with} you also paint\\
 %`I am painting with what you are painting with too.'
%\bg. Ik schilder met de kwast [waar -mee jij ook schildert].\\
 %I paint with the brush {where -with} you also paint\\
 %`I am painting with the brush that you are painting with too.'

The mismatching free relative in \ref{ex:grammaticalwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both construction \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

@@

I repeat the mismatching free relative in which \tit{waarmee} `met wat' appears in \ref{ex:mismatchwaarmee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental object. The instrumental object forms a proper constituent within the embedded clause, and the it can be realized as the \tsc{r}-pronoun and postposition \tit{waarmee} `with what'.\footnote{I assume that the accusative case requirement of \tit{gekocht} `bought' is satisfying by grafting a subconstituent of \tit{waarmee} `with what' \citep{bergsma2019}.}

@@

\ref{ex:summaryconst} summarizes what I showed in this section. \tit{Met wat} `with what' can never surface when \tit{met} `with' and \tit{wat} `what' form a proper constituent. It always becomes \tit{waarmee} `with what'. This is schematically shown in \ref{ex:waarmeefr}.
There are other contexts in which \tit{met wat} `with what' appears. This can be either when \tit{wat} `what' takes a complement, or when \tit{wat} is part of the a clause that \tit{met} `with' is not a part of. This last option is schematically showed in \ref{ex:metwatfr}, and it represents the mismatching free relative in \ref{ex:mismatchmetwat}.

\ex.\label{ex:summaryconst}
\a. [[met] [wat]] → [waarmee]\label{ex:waarmeefr}
\b. [met [wat [X]]]
\b. [met [[wat] [X]]]\label{ex:metwatfr}

\tit{Met} `with' is a preposition that combines with full DPs and animate pronouns. \tit{Wat} `what' is a wh-element that appears as subject or object.

In the next section I decompose \tit{waarmee} `with what' and \tit{met wat} `with what'. I show that both spell out the same set of features, but the distribution is different.

\section{Taking \tit{waarmee} and \tit{met wat} apart}

Along the way I introduce some background

Things to explain
*P pronoun-inanimate

\tit{r} is the locative
\tit{met} changes form

 The idea is that both expressions realize the same features, and \tit{waarmee} `with what' takes precedence when all features form a constituent. The morphemes I distinguish in \tit{waarmee} `with what' are \tit{w}, \tit{-aa-}, \tit{-r} and \tit{-mee}. Within \tit{met wat} `with what' I distinguish \tit{met}, \tit{w}, \tit{a} and \tit{-t}. \ref{ex:decompose} shows this as well.

\ex.\label{ex:decompose}
\a. w -aa -r -mee
\b. met w -a -t

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities and differences between the two forms. First, I identify \tit{w} and \tit{a} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{'rmee} `with it' and \tit{met 't} `with it'.

The elements \tit{w} and \tit{a} express the same syntactic structure in both expressions. The elements \tit{met} and \tit{'t} together also express the same features as \tit{-mee} and \tit{'r} together, but the distribution differs. \tit{Met} expresses less structure than \tit{-mee}, and \tit{'t} expresses more structure than \tit{'r}. \tit{Met} is shown to be an preposition, and \tit{-mee} is a postposition. It also becomes clear that the \tit{'r} in fact corresponds to the locative \tit{r} in Dutch.

@@

Let me start with the morphemes \tit{w} and \tit{a} that appear in both expressions. I assume that they correspond to the same syntactic structure in both \tit{waarmee} and \tit{met wat}. As I am interested in the differencs between the two expression, I do not discuss the featural content of \tit{w} and \tit{a} into depth.

@@

\ex. \begin{forest}
[WP
    [W, roof]
]
{\draw (.east) node[right]{⇔ \tit{w}}; }
\end{forest}\label{ex:entryw}

@@

\ex. \begin{forest}
[deixP
    [deix, roof]
]
{\draw (.east) node[right]{⇔ \tit{a}}; }
\end{forest}\label{ex:entrya}

@@

In this section I discuss the forms \tit{met 't} `with it` and \tit{'rmee} `with it'. I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following two observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a postposition, and its form changes into \tit{-mee}. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

@@

Let me start with the lexical entry for \tit{'t}. \tit{'t} `it' can be used as subject (associated with nominative), direct object (associated in accusative) and indirect object (associated with dative), as shown in \ref{ex:tsubobj}.

@@

 \ex.\label{ex:hemsubobj}
 \ag. Hij staat in de hal.\\
  \tsc{3sg.m.nom} stands in the hallway\\
  `He is standing in the hallway.'
 \bg. Ik zie hem.\\
  I see \tsc{3sg.m.acc}\\
  `I see it.'
 \bg. Ik heb hem een klap gegeven.\\
  I have \tsc{3sg.m.acc} a hit given\\ with both alternatives. I work the proposal out with't  realizing all cases.

For case, I follow \citet{caha2009} that case features case features are organized the containment relation in \ref{ex:casetree}. The higher, more complex cases contain the smaller, less complex cases.\footnote{there has been discussion on the genitive. starke on two accusatives and two datives. this paper: only small accusative and dative and genitive is above both of them.}

@@

 Following the distinctions from \citet{cardinaletti1996}, \tit{'t} `it' is a weak pronoun. It is not a clitic, because it can occur in sentence initial position, shown in \ref{ex:tnoclitic}. It is not a strong pronoun, because it cannot be coordinated, as indicated in  \ref{ex:tcoordinated}. \ref{ex:datcoordinated} shows that \tit{'t} `it' needs to combine with \tit{da-}/\tit{di-} to be able to be coordinated.

@@

 In other words, a lexically stored structure does not have to be identical to the syntactic structure. It is enough if the syntactic structure is contained in the lexically stored tree. This has a as consequence that the lexical entry in \ref{ex:entryt} can also be inserted in \ref{ex:tacc} and \ref{ex:tnom}.

@@

I follow \cite{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I left out here.}

@@

\ex. The Elsewhere Condition (\citealt{kiparsky1973}, formulated as in \citealt{caha2020}):\\
When two entries can spell out a given node, the more specific entry wins. Under the Superset Principle governed insertion, the more specific entry is the one which has fewer unused features

@@

\ref{ex:suspendedme} gives examples of the first person singular in a suspended affixation contexts. It shows that it is \tit{mæn} that appears as a caseless first conjunct and that the use of \tit{æz} is ungrammatical. This means that \tit{mæn} is the bare stem that combines with case markers, and \tit{æz} the suppletive nominative. In Section \ref{sec:derivation} I show how a derivation with this type of elements works in nanosyntax.

@@

\section{Conclusion and discussion}

>>>>

Dutch has the preposition \tit{met} `with' that expresses the instrumental. In \ref{ex:metdp}, \tit{met} `with' is combined with a full DP. The inanimate pronoun in Dutch is \tit{'t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{'t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{'t} `it' can be used as the object of a verb.

@@

\tit{Met} `with' and \tit{'t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{'r} `there' and the postposition \tit{mee} `with', as shown in \ref{ex:jarmee}.\footnote{The \tsc{r}-pronoun \tit{'r} in \ref{ex:jarmee} can be written as \tit{er}, \tit{der} and \tit{'r} and pronounced as respectively /ɛr/, /dər/ or /ər/. As far as I am aware, there is no clear meaning difference between these forms. See \citet{wesseling2018} for discussion.
In my examples I use \tit{'r}, but the other two forms fit just as well.}$^,$\footnote{In this paper I do not make any claims about the distinction between prefixes and prepositions, or suffixes and postpositions.}

@@

\tsc{R}-pronouns \citep{riemsdijk1978,koopman1994} are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r}. The adpositions they combine with obligatorily follow the \tsc{r}-pronoun, see \ref{ex:jarmee} and \ref{ex:neemeer}.
Notice also that the preposition \tit{met} `with' differs phonologically from the postposition \tit{mee} `with` (see \ref{ex:metdp} and \ref{ex:jarmee}).

@@

The main question I address in this chapter is how to correctly rule out \tit{met 't} `with it' in \ref{ex:neemett}, and let the \tsc{r}-pronoun and postposition in \ref{ex:jarmee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout mechanisms. Just like \citet{riemsdijk1978} I analyze \tsc{r}-pronouns and postpositions as a type of allomorph of the preposition and pronoun. The crucial difference in the current approach is that spellout rules out the ungrammaticality of the preposition and pronoun, and not the stipulation of a filter.

This chapter is an in-depth study of the instrumental \tsc{r}-pronoun and postposition \tit{'r -mee} `with it' and \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative is intuitive, as many prepositions express locations, directions etc. However, an instrumental expresses an instrument, which does not have a meaning component assosicated with location. Ideally, an analysis treats the syncretism not as accidental but allows for the locative meaning to be absent.
The second reason why I focus on this particular \tsc{r}-pronoun and postposition has to do with the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun. This last observervation has so far remained unexplained.

The main generalization is that the instrumental \tsc{r}-pronoun and postposition \tit{waarmee} `with what' takes precedence over the instrumental preposition and inanimate pronoun \tit{met wat} `with what'. This generalization is subject to an important condition: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition and pronoun appear. This can straightforwardly follow in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work this idea out capturing the following observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, regular pronouns appear with prepoositions, \tsc{r}-pronouns with postpositions. Third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I discuss the condition that \tsc{r}-pronouns and postpositions are subject too: they need to form a proper constituent. I show that \tsc{r}-pronouns and postpositions and prepositions and pronouns are in complementary distribution, depending on whether all features form a proper constituent or not.
In Section \ref{sec:takingapart} I decompose \tit{waarmee} `with what' and \tit{met wat} `with what', and I connect them to parts of syntactic structure. Section \ref{sec:derivation} shows in a derivation how \tit{waarmee} `with what' is derived and the role that constituency plays. Section \ref{sec:conclusion} concludes and discusses possible extensions of the analysis. All examples in this paper are from Dutch, unless indicated otherwise. Unmarked examples are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all features that are to be spelled out form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{'rmee} `with it' and \tit{waarmee} `with what` are the default, that normally they appear instead of \tit{met 't} `with it' and \tit{met wat} `with what`. Then I show an exception, in which \tit{met wat} `with what` has to be used and \tit{waarmee} `with what` cannot. I argue that the crucial difference between the exception and other cases is that in the exception the features do not form a proper constituent.

@@

In what follows I discuss the distribution of \tsc{r}-pronouns and regular pronouns in general \citep{riemsdijk1978,koopman2000}. I start with the personal pronouns and then return to the wh-pronouns.

Dutch has the accusative personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{het} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

@@

The example in \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. Repeating from the introduction, the inanimate personal pronoun \tit{het} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an \tsc{r}-pronoun appears. This is illustrated in \ref{ex:preper}. \ref{ex:erprep} shows that the \tsc{r}-pronoun obligatorily has moved to the left of the adposition.

@@

\tit{Met} is not the only preposition with which this happens. For example, \tit{op} `on' and \tit{in} `in' are two prepositions that also do not combine with the inanimate personal pronoun \tit{'t}, but the \tsc{r}-pronoun and postposition is used obligatorily.

@@

The situation of the inanimate wh-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as an object of a verb (see \ref{ex:wat}), but not as an object of a preposition \ref{ex:metwat}. In that case, the \tsc{r}-pronoun and postposition \tit{waarmee} `with what' appears, as shown in \ref{ex:waarmee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

@@

In sum, \tit{'t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by respectively \tit{'r} `there' and \tit{waar} `where'.

@@

This section discusses an exception to what is shown in the previous section. I show an instance in which \tit{met wat} `with what' has to be used instead of \tit{waarmee} `with what'. This instance comes from a mismatching free relative construction.

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I give an example in which the \tsc{r}-pronoun and postposition appears before I get to the exception. Consider \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun and postposition \tit{waarmee} `with what' is used here. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.\footnote{In this example, \tit{waar} `where' takes \tit{-mee} `with' to the left edge of the embedded clause. It is also possible for \tit{-mee} `with' to be stranded, and \tit{waar} `where' to be moved to the left edge of the embedded clause on its own.

@@

\ex.
\ag. Ik heb gekocht waarmee jij schildert.\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

If the predicates are switched around between the clauses, the \tsc{r}-pronoun and postposition do not appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental object in the main clause and \tit{gekocht} `bought' combines with an accusative object in the embedded clause. The use of an \tsc{r}-pronoun and postposition is ungrammatical, as indicated by \ref{ex:schilderwaarmee}. Instead, a combination of the regular instrumental preposition \tit{met} `with' and the regular wh-pronoun \tit{wat} `what' in used.

@@

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the embedded clause predicate with an instrumental, \tit{waarmee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the embedded clause predicate with an accusative, \tit{waarmee} `with what' is ungrammatical and \tit{met wat} `with what' is used.

@@

In the remainder of this section I argue that the crucial point of \ref{ex:schildermetwat} is that the instrumental object does not form a proper constituent, i.e. it is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns and postpositions contain an instrumental object that does form a proper constituent.

@@

In each of these examples the instrumental object forms a proper constituent at a certain point in the derivation. In \ref{ex:const1}, the instrumental object forms a proper constituent in the surface order, as shown in \ref{ex:const1stage}. In \ref{ex:const2}, the instrumental object forms a proper constituent before wh- and V2- movement, shown in \ref{ex:const2stage}.
The structure in \ref{ex:const3stage} represents a stage in the derivation of the embedded clauses in \ref{ex:const3} and \ref{ex:const4}. Again, in the stage, which comes before relative movement of the pronoun to the left periphery of the relative clause, the instrumental object forms a proper constituent.

@@

@@

I repeat the mismatching free relative in which \tit{waarmee} `with what' appears in \ref{ex:mismatchwaarmee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental object. The instrumental object forms a proper constituent within the embedded clause, and the it can be realized as the \tsc{r}-pronoun and postposition \tit{waarmee} `with what'.\footnote{I assume that the accusative case requirement of \tit{gekocht} `bought' is satisfying by grafting a subconstituent of \tit{waarmee} `with what' \citep{bergsma2019}.}

@@

 The mismatching free relative in \ref{ex:grammaticalwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
 In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both constructions \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

\ref{ex:summaryconst} summarizes what I showed in this section. \tit{Met wat} `with what' can never surface when \tit{met} `with' and \tit{wat} `what' form a proper constituent. It always becomes \tit{waarmee} `with what', as shown in \ref{ex:waarmeefr}.

\ex.\label{ex:summaryconst}
\a. [[met] [wat]] → [waarmee]\label{ex:waarmeefr}
\b. [met [wat [X]]]\label{ex:metwatx}
\b. [met [[wat] [X]]]\label{ex:metwatfr}

\tit{Met wat} `with what' can appear when the instrumental object does not form a proper constituent. This can be either when \tit{wat} `what' takes a complement before it combines with \tit{met} `with', as shown in \ref{ex:metwatx}, representing the examples in \ref{ex:moremetwat}.  The other option is that \tit{wat} `what' is part of the a clause that \tit{met} `with' is not a part of, schematitized as \ref{ex:metwatfr} as an illustration of \ref{ex:mismatchmetwat}.

\section{Decomposing \tit{waarmee} and \tit{met wat}}\label{sec:takingapart}

In this section I decompose \tit{waarmee} `with what' and \tit{met wat} `met wat' into smaller units, and I connect them to parts of syntactic structure. Both spell out the same set of features but the distribution is different. I decompose \tit{waarmee} and \tit{met wat} as I show in \ref{ex:decompose}. Earlier works have decomposed these pronouns in similar ways (cf. \citet{hachem2015,noonan2017dutch,wesseling2018}).

\ex.\label{ex:decompose}
\ag. w -aa -r -mee\\
\tsc{w} -\tsc{deix} -\tsc{loc} -\tsc{ins}\\
\bg. met w -a -t\\
\tsc{ins} \tsc{w} -\tsc{deix} -\tsc{n.sg}\\
\z.
\z.

In this section I first identify \tit{w} and \tit{a} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{'rmee} `with it' and \tit{met 't} `with it'. Along the way I introduce some necessary theoretical background on Nanosyntax.

@@

Let me start with the morphemes \tit{w} and \tit{a} that appear in both expressions. I assume that they correspond to the same syntactic structure in both \tit{waarmee} `with what' and \tit{met wat} `with what'. As I am interested in the differencs between the two expression, I do not discuss the featural content of \tit{w} and \tit{a} into depth.

@@

\ex. \begin{forest}
[\tsc{wP}
    [W, roof]
]
{\draw (.east) node[right]{⇔ \tit{w}}; }
\end{forest}\label{ex:entryw}

@@

\ex. \begin{forest}
[\tsc{deixP}
    [\tsc{deix}, roof]
]
{\draw (.east) node[right]{⇔ \tit{a}}; }
\end{forest}\label{ex:entrya}

@@

In this section I discuss the forms \tit{met 't} `with it` and \tit{'rmee} `with it'. I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following three observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a postposition. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{met} changes form into \tit{-mee}. Third, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

@@

Let me start with the lexical entry for \tit{'t}. \tit{'t} `it' can be used as a subject (associated with nominative), direct object (associated in accusative) and indirect object (associated with dative), as shown in \ref{ex:tsubobj}.

@@

 \ex.\label{ex:hemsubobj}
 \ag. Hij staat in de hal.\\
  \tsc{3sg.m.nom} stands in the hallway\\
  `He is standing in the hallway.'
 \bg. Ik zie hem.\\
  I see \tsc{3sg.m.acc}\\
  `I see it.'
 \bg. Ik heb hem een klap gegeven.\\
  I have \tsc{3sg.m.dat} a hit given\\
  `I gave him a hit.'

For case, I follow \citet{caha2009} that case features are organized as in the containment relation in \ref{ex:casetree}. The higher, more complex cases contain the smaller, less complex cases.\footnote{I follow \citet{starke2017} in that there are two types of accusatives and datives. In this paper I only include the S-\tsc{acc} and S-\tsc{dat}, which are placed below the \tsc{gen}.}

@@

 Following the distinctions from \citet{cardinaletti1996}, I assume \tit{'t} `it' is a weak pronoun. It is not a clitic, because it can occur in sentence initial position, shown in \ref{ex:tnoclitic}. It is not a strong pronoun, because it cannot be coordinated, as indicated in  \ref{ex:tcoordinated}. \ref{ex:datcoordinated} shows that \tit{'t} `it' needs to combine with \tit{da-}/\tit{di-} to be able to be coordinated.

@@

 In other words, a lexically stored structure does not have to be identical to the syntactic structure. It is enough if the syntactic structure is contained in the lexically stored tree. This has as a consequence that the lexical entry in \ref{ex:entryt} can also be inserted in \ref{ex:tacc} and \ref{ex:tnom}.

@@

I follow \citet{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I have left out here.}

@@

\ex. The Elsewhere Condition (\citealt{kiparsky1973}, formulated as in \citealt{caha2020}):\\
When two entries can spell out a given node, the more specific entry wins. Under the Superset Principle governed insertion, the more specific entry is the one which has fewer unused features.

@@

\ref{ex:suspendedme} gives examples of the first person singular in a suspended affixation contexts. It shows that it is \tit{mæn} that appears as a caseless first conjunct and that the use of \tit{æz} is ungrammatical. This means that \tit{mæn} is the bare stem that combines with case markers, and \tit{æz} the suppletive nominative. In Section \ref{sec:derivation} I show how a derivation with this type of elements works in Nanosyntax.

@@

\section{Conclusion and discussion}\label{sec:conclusion}
